---
title: "Liberia Community-based Management of Acute Malnutrition Coverage Assessment"
subtitle: "A report prepared for the Liberia Ministry of Health and UNICEF Liberia"
author: 
    - name: MSc in International Health and Tropical Medicine Class 2023
    - name: University of Oxford
date: '`r Sys.Date()`'
output: 
  html_document:
    toc: true
    toc_depth: 3
    self_contained: true
    code_download: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  error = FALSE,
  message = FALSE,
  echo = FALSE
)

## Load libraries
library(dplyr)
library(gtsummary)

## Load functions in R folder
for (f in list.files("R", full.names = TRUE)) source (f)

## Get data
nut <- get_data()
```

## Introduction

A three-year nutrition programme has been implemented in Liberia by the Ministry of Health and UNICEF aimed at tackling child undernutrition in the country. Funded by [Power of Nutrition](http://www.powerofnutrition.org) and [UNICEF UK](https://www.unicef.org.uk), the programme has been implemented across 15 counties in Liberia starting from January 2017 up to December 2019. The overall aim of the programme is to improve the coverage of direct nutrition interventions or what is commonly termed nutrition-specific interventions, i.e. interventions or programmes that address the immediate determinants of foetal and child nutrition and development — adequate food and nutrient intake, feeding, caregiving and parenting practices, and low burden of infectious diseases. The current programme supports the following specific key interventions: 1) *treatment of severe acute malnutrition (SAM) within the community-based management of acute malnutrition (CMAM) programme for children 6-59 months*; 2) *vitamin A supplementation for children 6-59 months*; 3) *promotion of appropriate infant and young child feeding (IYCF) practices among pregnant or lactating women*; 4) *multiple micronutrient powder (MNP) supplementation for children 6-23 months*; and, 5) *iron and folic acid (IFA) supplementation for pregnant women*.

To assess the programme's progress towards its overall aim, two coverage assessments have been implemented - the first at the halfway point of the programme and the second at the end. Only two programme areas were selected for the assessments: *Urban Montserrado (Greater Monrovia)* district and *Grand Bassa* county.

This document reports on the results of the coverage assessment of the CMAM programme for treating SAM children 6-59 months old.

## Description of the sample
<!--- INSERT TEXT AND LITERATE CODE BELOW THIS LINE --->





## Nutrition screening coverage
<!--- INSERT TEXT AND LITERATE CODE BELOW THIS LINE --->

The following data processing was first performed on the data before continuing on with the analysis.

```{r process_data_screen, echo = TRUE}
## Subset data: Transform variables of interest into table nut1 
nut1 <- nut %>% select(survey_round,county,sex,muac_screen,oedema_screen)

## Clean data in subset: Change 99 to NA 
### Clean data only after having made subset nut1. 
### Code don't work if data is cleaned in nut and then subset. ???
nut1$muac_screen <- ifelse(nut1$muac_screen == 99, NA, nut1$muac_screen)
nut1$muac_screen[nut$muac_screen == 1] <- "Y"
nut1$muac_screen[nut$muac_screen == 2] <- "N"

nut1$oedema_screen <- ifelse(nut1$oedema_screen == 99, NA, nut1$oedema_screen)
nut1$oedema_screen[nut$oedema_screen == 1] <- "Y"
nut1$oedema_screen[nut$oedema_screen == 2] <- "N"

nut1$sex <- ifelse(nut1$sex == 1, "Male", "Female")
```

### MUAC screening coverage

The following table presents MUAC screening coverage by survey round.

```{r muac_screen_table, echo = TRUE}
nut1 %>% 
  tbl_summary(
    by = survey_round, 
    include = muac_screen,
    statistic = list(all_continuous() ~ "{mean} ({sd})")
  )
```

The endline data shows LITTLE if NO improvement in MUAC screening coverage among the sample children vs. baseline. Percentage of children had MUAC screened (answered 1 to survey) stayed relatively the same from baseline to endline. The percentage of children without MUAC screening went up from 73% to 93%, mainly due to children from the NA population at baseline moving over. This may suggest a better awareness in this population related to not having gotten MUAC screening (?).

The following table presents MUAC screening coverage by survey round and by sex of the child.

```{r muac_screen_by_sex_table, echo = TRUE}
nut1 %>% 
  tbl_strata(
    strata = sex, 
    .tbl_fun = ~ .x %>% 
      tbl_summary(
        by = survey_round,
        include = muac_screen,
        statistic = list(all_continuous() ~ "{mean} ({sd})")
      )
  )
```

There is no apparent difference in MUAC screening coverage between girls and boys in the sample. Or, gender doesn't seem to influence MUAC screening in children.

The following table presents MUAC screening coverage by survey round and by survey location.

```{r muac_screen_by_location_table, echo = TRUE}
nut1 %>% 
  tbl_strata(
    strata=county,
    .tbl_fun = ~ .x %>% 
      tbl_summary(
        by = survey_round,
        include = muac_screen,
        statistic = list(all_continuous() ~ "{mean} ({sd})")
      )
   )
```

The data suggests an increase in children receiving MUAC screening at endline in Grand Bassa vs. in Urban Montserrado. In fact, it seems that the coverage of MUAC screening in Urban Montserrado decreases (from 4.5% at baseline to 3% at endline). This suggests Grand Bassa to have done a better job in MUAC screening than Montserrado. However, neither location has a significant improvement in MUAC screening coverage. The overall trend of increasing in no screening is observed here but this is again due to population who answered NA at baseline moving to answered no to screening at endline.

### Oedema screening coverage

The following table presents oedema screening by survey round.

```{r oedema_screen_table, echo = TRUE}
nut1 %>% 
  tbl_summary(
    by = survey_round, 
    include = oedema_screen,
    statistic = list(all_continuous() ~ "{mean} ({sd})")
  )
```

Similar to MUAC screening, the endline data shows little if no improvement in oedema screening coverage among the sample children vs. baseline. The same observation applies for movement of populations who answered NA for oedema screening at baseline to answering no to Oedema screening at endline. This may suggest again a better sense of awareness and may serve as a reflection of public engagement working?

The following table presents oedema screening by survey round and by sex.

```{r oedema_screen_by_sex, echo = TRUE}
nut1 %>% 
  tbl_strata(
    strata = sex,
    .tbl_fun = ~ .x %>% 
      tbl_summary(
        by = survey_round,
        include = oedema_screen, 
        statistic = list(all_continuous() ~ "{mean} ({sd})")
      )
  )
```

The data suggests oedema screening improves SLIGHTLY in females vs. in males. However, this difference does not seem big enough to call for gender as a determinant in improvement of oedema screening. p-value may be needed to determine if the difference is significant.

The following table presents oedema screening by survey round and by survey location.

```{r oedema_screen_by_location, echo = TRUE}
nut1 %>% 
  tbl_strata(
    strata = county,
    .tbl_fun = ~ .x %>% 
      tbl_summary(
        by = survey_round,
        include = oedema_screen,
        statistic = list(all_continuous() ~ "{mean} ({sd})")
      )
  )
```

The same trend is observed here for oedema screening. The data suggests an increase in oedema screening coverage in Grand Bassa but a decrease in Urban Montserrado, suggesting Grand Bassa doing a better job. 

The following table presents all the variables of interest with *p-values* for *chi-squared test* comparing results for each category.

```{r summary_table, echo = TRUE}
nut1 %>% 
  tbl_summary(
    by = survey_round, 
    statistic = list(all_continuous() ~ "{mean} ({sd})")
  ) %>% 
  add_p() 
```

*Pearson's Chi-squared test* at the end help to show that difference between endline and baseline is indeed statistically significant. But this may be driven mainly by the number answering no to screening increasing between baseline and endline, not a reflection of how well the screening has increased between the 2 timepoints.

## Prevalence of acute undernutrition
<!--- INSERT TEXT AND LITERATE CODE BELOW THIS LINE --->





## Coverage of SAM treatment
<!--- INSERT TEXT AND LITERATE CODE BELOW THIS LINE --->

### Number of SAM cases found

The following table presents the number of cases of SAM found in the overall sample.

```{r sam_counts_table, echo = TRUE}
nut$sam[nut$age >= 6 & nut$age < 60 & (nut$muac >= 11.5 | nut$oedema == 2)] <- 'no'
nut$sam[nut$age >= 6 & nut$age < 60 & (nut$muac < 11.5 | nut$oedema == 1)] <-'yes'

sam_table <- table(nut$sam)

sam_table
```

There were a total of **`r sam_table[2]` SAM cases** found in the overall sample.

The following table presents the number of SAM cases found in the sample by location.

```{r sam_counts_location, echo = TRUE}
sam_location_table <- table(nut$sam, nut$county)

sam_location_table
```

There were a total of **`r sam_location_table[2, 1]` SAM cases** found in Grand Bassa compared to **`r sam_location_table[2, 2]` SAM cases** found in Urban Montserrado.

The following table presents the number of SAM cases found in the sample by survey round.

```{r sam_counts_round, echo = TRUE}
sam_rounds_table <- table(nut$sam, nut$survey_round)

sam_rounds_table
```

There were a total of **`r sam_rounds_table[2, 1]` SAM cases** found at baseline survey round compared to **`r sam_rounds_table[2, 2]` SAM cases** found at endline survey round.

The following table prsents the number of SAM cases found in the sample by location and survey round.

```{r sam_location_round, echo = TRUE}
sam_location_rounds_table <- table(nut$sam, nut$survey_round, nut$county)

sam_location_rounds_table
```

In Grand Bassa, there were a total of **`r sam_location_rounds_table[2, 1, 1]` SAM cases** found at baseline and **`r sam_location_rounds_table[2, 2, 1]`** SAM cases found at endline.

In Urban Montserrado, there were a total of **`r sam_location_rounds_table[2, 1, 2]` SAM cases** found at baseline and **`r sam_location_rounds_table[2, 2, 2]`** SAM cases found at endline.

### Case-finding effectiveness

The following table presents the case-finding effectiveness of the CMAM programme overall.

```{r case-finding}
samcov <- table(nut$sam, nut$cov_status)

samcov
## prop.table(samcov) * 100      This code gives proportion of SAM cases in the
                              ## programme by all children in the sample which 
                              ## is not consistent with the coverage estimator

## This code calculates the proportions for only those children who are SAM and 
## in the programme
samcovprop <- prop.table(samcov[2, ] * 100)

samcovprop
```

The case-finding effectiveness of the CMAM programme overall was at **`r round(samcovprop[1], digits = 2)`%**. 

The following table presents the case-finding effectiveness of the CMAM programme by location.

```{r case-finding-location, echo = TRUE}
samcovloc <- table(nut$sam, nut$cov_status, nut$county)

samcovloc

## prop.table(samcovloc) * 100     This code gives proportion of SAM cases in the
                                ## programme by all children in the sample
                                ## which is not consistent with the coverage
                                ## estimator

## This code calculates the proportions for only those children who are SAM
## and in the programme for each location
samcov_grandbassa <- prop.table(samcovloc[2, , 1]) * 100

samcov_urbanmontserrado <- prop.table(samcovloc[2, , 2]) * 100
```

In Grand Bassa, the case-finding effectiveness of the CMAM programme was at **`r round(samcov_grandbassa[1], digits = 2)`%**. In Urban Montserrado, the case-finding effectiveness of the CMAM programme was at **`r round(samcov_urbanmontserrado[1], digits = 2)`%**.

